---
title: "Playing and Exporting Coordinate Animations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Playing and Exporting Coordinate Animations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", out.width = "100%")
library(ivue)
have.rgl <- nzchar(system.file(package = "rgl"))
have.magick <- requireNamespace("magick", quietly = TRUE)
```

`animate.frames()` turns recorded coordinates into a browser player.
`write.animation.gif()` exports the same retained frames to a GIF for a
README, presentation, or website. `ivue` draws the frames; the package or
simulation producing them remains responsible for their computation.

Playback requires `rgl`. GIF export additionally requires `magick`:

```{r installation, eval=FALSE}
install.packages(c("rgl", "magick"))
```

## Watch a Sierpinski triangle unfold

This small example is a recorded two-dimensional GRIP layout of a level-4
Sierpinski triangle. Forty frames were selected across a longer solve. The
coordinates are unchanged: no frame was aligned to a target shape, rotated,
or normalized for presentation. The package includes the record so this
vignette does not require `grip` to run.

```{r triangle-data}
triangle <- readRDS(system.file("extdata", "sierpinski-trace.rds", package = "ivue"))
length(triangle$frames)
dim(triangle$frames[[1]])
head(triangle$meta[, c("phase", "active_vertices")])
```

```{r triangle-player, eval=have.rgl}
triangle.labels <- paste0("Frame ", triangle$original.frame.index,
                           " | ", triangle$meta$phase,
                           " | ", triangle$meta$active_vertices, " vertices")
triangle.player <- animate.frames(
  triangle$frames, edges = triangle$edges, labels = triangle.labels,
  fps = 5, col = "#C24E25", point.size = 4,
  edge.col = "#314E6ECC", edge.width = 1,
  background.color = "#FAF7F0", height = 450
)
triangle.player
```

Press **Play** to start; the button becomes **Pause**. Drag the slider to
inspect a particular frame. With the slider focused, use the arrow keys to
move one frame at a time. **Slower** and **Faster** change playback speed,
**Reverse** changes direction, and **Reset** returns to the beginning.
Drag inside the scene to rotate it, including during playback. The default
2D camera looks straight down onto the xy plane.

The recorded frames all have the same number of rows. A vertex that has not
yet been introduced has an entirely missing (`NA` or `NaN`) row. Its point and all incident
edges stay hidden until its coordinates are finite. Rows remain associated
with the same vertices throughout playback; they must not be reordered.

## Supply your own frames

Frames are a list of numeric matrices with two or three columns. Every
matrix must have identical dimensions. If row names are supplied, they must
be unique and identical across frames. A row must be completely finite or
completely missing (`NA` or `NaN`); partial missing coordinates and infinity
are errors. GRIP records its inactive vertices using `NaN` rows.

Here a triangle gains its third vertex and then expands. The second edge
appears only when its missing endpoint is introduced.

```{r small-example, eval=have.rgl}
X <- rbind(a = c(0, 0), b = c(1, 0), c = c(0.5, 0.9))
first <- X
first[3, ] <- NA
small <- animate.frames(
  list(first, X, X * 1.4),
  edges = rbind(c(1, 2), c(2, 3), c(3, 1)),
  labels = c("Two vertices", "Triangle", "Expanded triangle"),
  fps = 1, loop = FALSE, point.size = 9, height = 280
)
small
```

Vertices may disappear as well as appear. An entirely empty frame is
allowed, provided at least one frame contains a finite point. The stricter
finite-coordinate requirement of ordinary `plot3D` functions is unchanged.

## Animate a changing 3D surface

Any sequence of recorded coordinates can be played. This example explicitly
constructs a sequence from a plane to a saddle; it is an illustrative
deformation, not the output of a layout optimizer. The xy positions and
vertex identities are fixed while the height changes.

```{r saddle-frames}
side <- 9L
grid <- expand.grid(x = seq(-1, 1, length.out = side),
                    y = seq(-1, 1, length.out = side))
amplitudes <- seq(0, 1.2, length.out = 25)
saddle.frames <- lapply(amplitudes, function(a) {
  cbind(x = grid$x, y = grid$y, z = a * (grid$x^2 - grid$y^2))
})
ids <- matrix(seq_len(nrow(grid)), side, side)
saddle.edges <- rbind(
  cbind(as.vector(ids[-side, ]), as.vector(ids[-1, ])),
  cbind(as.vector(ids[, -side]), as.vector(ids[, -1]))
)
height.scale <- color.scale.cont(saddle.frames[[25]][, 3], center = 0,
                                 palette = c("#2455A4", "#ECE6C2", "#B83232"))
point.colors <- map.colors(saddle.frames[[25]][, 3], height.scale)$colors
```

```{r saddle-player, eval=have.rgl}
saddle.player <- animate.frames(
  saddle.frames, edges = saddle.edges,
  labels = sprintf("Saddle amplitude = %.2f", amplitudes),
  col = point.colors, point.size = 6, edge.col = "#314E6E99",
  camera = camera.zup(elevation = 25, turn = -130),
  fps = 8, height = 450
)
saddle.player
```

Colors describe each vertex's **final height** and remain fixed across
frames. This makes identity easy to follow; they do not encode instantaneous
height. Frame-dependent colors and animated mesh faces are outside this
initial API. `edges` draws a wire grid without filling its faces.

## Export interactive HTML

The returned object is an ordinary htmlwidget. Saving it preserves the
player and camera controls and does not require Shiny or a running R session
for playback. No file is saved merely by constructing a player.

```{r html-export, eval=FALSE}
htmlwidgets::saveWidget(triangle.player, "triangle-playback.html", selfcontained = TRUE)
htmlwidgets::saveWidget(saddle.player, "saddle-playback.html", selfcontained = TRUE)
```

Self-contained HTML requires Pandoc and bundles widget assets into one file.
Use `selfcontained = FALSE` to save the HTML alongside a dependency folder;
keep that folder with the HTML when sharing it.

## Export a GIF

GIF export uses the frames retained in the player, including inactive
vertices and edges. This example writes into R's temporary directory, leaving
your working directory untouched. Supply your own destination to keep it.

```{r gif-export, eval=have.rgl && have.magick}
gif.path <- tempfile(fileext = ".gif")
write.animation.gif(triangle.player, gif.path, fps = 5,
                     width = 480, height = 480, final.hold = 2)
knitr::include_graphics(gif.path)
```

```{r keep-gif, eval=FALSE}
write.animation.gif(saddle.player, "saddle.gif", fps = 8,
                     width = 600, height = 500, final.hold = 2,
                     loop = TRUE, overwrite = FALSE)
```

The GIF has the widget's **initial camera orientation**, not a rotation made
later in the browser. Set `camera` explicitly when constructing the player to
choose the exported view. Export requires an orthographic camera (`fov = 0`),
which is the animation default. It renders a diagram from the recorded
coordinates using R graphics and `magick`; it is not a screenshot of the
WebGL scene. Point appearance can differ slightly. Edges are painted before
points, so complex 3D intersections do not have WebGL depth-buffer semantics.

The last frame has `final.hold` additional seconds per loop. GIF frame delays
are rounded to centiseconds. Existing files are protected unless you set
`overwrite = TRUE`.

## Control size and preserve diagnostic meaning

Playback changes only which recorded frame is displayed. It does not
interpolate missing frames, align to a target, or recenter each frame. All
original frames determine one fixed viewing box, so translation, contraction,
and expansion remain visible. Rotating or zooming the camera changes the
view, not the stored coordinates.

By default, at most 100 evenly spaced frames are retained, including the
first and last. If this limit is exceeded, a message reports the subsampling.
For a diagnostic inspection, choose frames explicitly or retain all of them:

```{r frame-selection, eval=have.rgl}
selected <- animate.frames(triangle$frames, triangle$edges,
                            frame.index = c(1, 10, 20, 30, 40),
                            labels = triangle.labels, fps = 2)
attr(selected, "ivue.animation")$frame.index
```

```{r all-frames, eval=FALSE}
all.frames <- animate.frames(my.frames, edges = my.edges, max.frames = NULL)
```

Every retained frame has equal duration. A subsampled animation therefore
shows the order of the solve, not elapsed computation time. Widget size grows
with both frame count and graph size; large traces benefit from an explicit
selection. The exported GIF uses that same selection.

## Use a fresh GRIP trace

With a recent `grip` installation, pass its recorded matrices directly.
This optional recipe is not evaluated when building the vignette:

```{r fresh-grip, eval=FALSE}
edges <- grip::edges.sierpinski.triangle(level = 4)
tr <- grip::trace.grip(edges, n = max(edges), dim = 2,
                       preset = "carpet", seed = 1,
                       trace = "round", trace.every = 1)
player <- ivue::animate.frames(
  tr$frames, edges,
  labels = paste(tr$meta$phase, "|", tr$meta$active_vertices, "vertices"),
  max.frames = 80, fps = 6
)
player
ivue::write.animation.gif(player, "grip-trace.gif")
```

The same interface accepts three-dimensional traces and frames recorded by
other solvers. Neither `animate.frames()` nor GIF export calls `grip` or
computes a layout.

The bundled triangle's package version, source commit, settings, and selected
indices are recorded alongside its coordinates:

```{r provenance}
triangle$provenance
```

Its generation script is `tools/generate-animation-example.R` in the ivue
source repository. The stored record allows the vignette to keep working
independently of future changes to GRIP's layout algorithm.
