---
title: "Example data and reusable visualization recipes"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Example data and reusable visualization recipes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3)
library(ivue)
have.rgl <- nzchar(system.file(package = "rgl"))
```

A reusable ivue example starts with coordinates, observation IDs, annotations,
and the meaning of any connections or frames. This guide constructs three
small illustrative inputs in base R; it does not fit an embedding or propose
a dataset-generation API. Use the [function guide](function-guide.html) to
choose an entry point and the [introduction](ivue-introduction.html) for
extended plotting controls.

## Choose a recipe

| What you need | Functions to combine | What to retain or share |
|---|---|---|
| Finite n-by-3 positions and one numerical value per row | `color.scale.cont()`, `plot3D.cont()` | Coordinates, IDs, scale, initial camera, HTML widget. |
| The same positions and category labels | `color.scale.groups()`, `plot3D.groups()` | Named palette, level order, annotations, HTML widget. |
| A vertex set, weighted edges, and supplied positions | `prepare.graph()`, `plot3D.graph()`, optional layers | Vertex IDs, normalized edge order, weight meaning, coordinates. |
| Triangle indices or an independent height grid | `layer3D.mesh()` or `layer3D.surface()` with a point plot | Connectivity or grid coordinates; these have different reuse semantics. |
| Stable observation rows over two or more frames | `map.colors()`, `animate.frames()` | Original frames and labels, retained indices, fixed colors, player. |
| A finished static or animated view | `htmlwidgets::saveWidget()`; for animation, `write.animation.gif()` | Interactive HTML plus dependencies, or a noninteractive GIF. |

Data, scales, cameras, and layer specifications can be kept together in an R
list and saved with base R's `saveRDS()`. Keep the original scientific inputs
and preparation choices as well as the widget.

## Construct a point cloud

This helix samples 24 equally spaced parameter values from one complete turn.
Height is a numerical annotation; the sign of height defines two illustrative
categories. These categories are assigned by construction, not discovered
clusters. The first and last points share horizontal position but differ in
height, so they are distinct observations.

```{r point-data}
n <- 24L
t <- seq(0, 2 * pi, length.out = n)
X <- cbind(x = cos(t), y = sin(t), z = seq(-1, 1, length.out = n))
ids <- sprintf("point-%02d", seq_len(n))
rownames(X) <- ids
annotations <- data.frame(
  id = ids, height = X[, "z"],
  half = factor(ifelse(X[, "z"] < 0, "Lower", "Upper"),
                levels = c("Lower", "Upper"))
)
# Simulate an annotation table arriving in another order, then align it.
annotations <- annotations[rev(seq_len(n)), ]
stopifnot(!anyNA(annotations$id), !anyDuplicated(annotations$id),
          setequal(annotations$id, rownames(X)))
annotations <- annotations[match(rownames(X), annotations$id), ]
stopifnot(identical(annotations$id, rownames(X)),
          identical(dim(X), c(n, 3L)), all(is.finite(X)))
head(annotations, 3)
```

This explicit matching matters: ordinary point plots pair annotations by row
position, not by vector names. If you subset or reorder observations, apply
the same operation to coordinates and annotations and rebuild any row-indexed
connectivity. Missing positions are errors in static plots. Missing annotation
values can remain and receive the scale's missing color.

## Share a scale and camera

Compare height before and after an illustrative change to the annotation,
while keeping positions and row identity fixed. The second value is half the
first; it is not another embedding. A single numerical scale makes that
reduction visible. Separate automatic scales would color both panels almost
identically despite their different values.

```{r shared-settings}
original <- annotations$height
reduced <- original / 2
scale <- color.scale.cont(c(original, reduced), limits = c(-1, 1), center = 0,
                          palette = c("#2166AC", "#F7F7F7", "#B2182B"))
camera <- camera.zup(elevation = 20, turn = -135, fov = 0, zoom = 0.65)
original.mapping <- map.colors(original, scale)
reduced.mapping <- map.colors(reduced, scale)
stopifnot(identical(original.mapping$legend, reduced.mapping$legend))
original.mapping$legend
```

```{r shared-views, eval=have.rgl}
original.view <- plot3D.cont(X, original, scale = scale, camera = camera,
  aspect = "equal", point.size = 7, height = 360,
  legend.title = "Original height", legend.width = 120)
reduced.view <- plot3D.cont(X, reduced, scale = scale, camera = camera,
  aspect = "equal", point.size = 7, height = 360,
  legend.title = "Half height", legend.width = 120)
original.view
reduced.view
```

Both widgets start with identical coordinates, IDs, point sizes, aspect,
projection, camera, and scale limits. Colors alone encode the changed values;
no coordinate scaling or alignment occurs. At the negative endpoint the
original view is dark blue, while half height has a lighter blue. Each widget
rotates independently: this workflow does **not** synchronize interaction.
For a noninteractive reader, the following small table records the same change.

```{r shared-summary}
rows <- c(1L, 12L, 24L)
data.frame(id = ids[rows], original = original[rows], reduced = reduced[rows],
           original.color = original.mapping$colors[rows],
           reduced.color = reduced.mapping$colors[rows])
```

A categorical alternative uses the same observation order. This constructs
and checks another view without adding a third embedded widget to the guide.
Print `group.view` interactively to display it.

```{r group-recipe}
group.scale <- color.scale.groups(annotations$half,
                                  colors = c(Lower = "#2166AC", Upper = "#B2182B"))
map.colors(annotations$half, group.scale)$legend
```

```{r group-view, eval=have.rgl}
group.view <- plot3D.groups(X, annotations$half, scale = group.scale,
                            camera = camera, height = 300,
                            legend.width = 120, legend.title = "Helix half")
stopifnot(inherits(group.view, "htmlwidget"))
```

Equal colors across views establish a shared annotation mapping. Equal cameras
establish an initial orientation. Neither establishes equivalent geometries,
clustering quality, a developmental trajectory, or biological causation.

## Prepare a small embedded weighted graph

These four vertices and two connections are chosen for illustration. The
weights represent supplied target distances; they are not inferred from the
drawn segment lengths. The fourth vertex is isolated and is still retained.

```{r graph-data}
vertices <- c("start", "bend", "end", "isolate")
edges <- data.frame(from = c("start", "bend"), to = c("bend", "end"),
                    weight = c(2, 4))
coords <- rbind(start = c(0, 0, 0), bend = c(1, 1, 0),
                end = c(2, 0, 1), isolate = c(0, 2, 1))
graph <- prepare.graph(edges, vertices = vertices, weight.type = "distance")
stopifnot(identical(graph$vertices$id, vertices),
          nrow(graph$edges) == 2L, all(is.finite(coords)),
          identical(rownames(coords), graph$vertices$id))
graph$edges
```

The prepared edge endpoints are integer indices into `graph$vertices`, not
external IDs. We deliberately shuffle coordinate and annotation input order
below. Unlike a point plot, `plot3D.graph()` aligns these named inputs by exact
vertex ID. Layers still refer to the **prepared vertex order**.

```{r graph-view, eval=have.rgl}
coords <- coords[c("end", "start", "isolate", "bend"), , drop = FALSE]
values <- c(isolate = 40, bend = 20, start = 10, end = 30)
graph.view <- plot3D.graph(graph, X = coords, values = values,
  edge.width = c(1, 3), edge.col = "gray55",
  point.size = 8, legend.width = 100, height = 340,
  camera = camera.zup(zoom = 0.55),
  layers = list(layer3D.path(c(1, 3), col = "#D55E00", width = 2),
                layer3D.labels(1:4, vertices, offset = c(0, 0, 0.15))))
info <- attr(graph.view, "ivue")
stopifnot(identical(rownames(info$X), vertices),
          identical(info$mapping$colors,
                    map.colors(unname(values[vertices]), info$mapping$scale)$colors))
graph.view
```

The orange geometric path joins start directly to end without changing the
graph topology; the gray graph edges go through bend. Labels sit 0.15
coordinate units above their vertices. Edge widths are explicit visual choices in prepared edge
order, independent of weights. No layout is computed. See the
[introduction](ivue-introduction.html#weighted-graphs-and-geometric-layers)
for reciprocal adjacency lists, matrices, and optional igraph layouts.

## Record a short coordinate sequence

This three-frame triangle first gains a vertex and then expands by a factor
of 1.4 about the coordinate origin. The change is explicitly constructed;
these frames are neither solver iterations nor physical time measurements.

```{r frame-data}
triangle <- rbind(a = c(0, 0), b = c(1, 0), c = c(0.5, 0.9))
first <- triangle
first[3, ] <- NA_real_
frames <- list(first, triangle, triangle * 1.4)
frame.labels <- c("Two vertices", "Complete triangle", "Expanded by 1.4")
frame.edges <- rbind(c(1, 2), c(2, 3), c(3, 1))
stopifnot(length(frames) == length(frame.labels),
  all(vapply(frames, function(f) identical(dim(f), c(3L, 2L)) &&
    identical(rownames(f), rownames(triangle)) &&
    all(rowSums(is.finite(f)) == 2L | rowSums(is.na(f)) == 2L), logical(1))),
  all(frame.edges >= 1L & frame.edges <= nrow(triangle)))
```

```{r frame-view, eval=have.rgl}
player <- animate.frames(frames, edges = frame.edges, labels = frame.labels,
  fps = 1, loop = FALSE, max.frames = NULL,
  col = c("#2166AC", "#B2182B", "#009F87"), point.size = 9, height = 300)
player
```

Press **Play**, step with the slider, or use **Reset**. Blue and red are present
at the start; green and its two incident edges appear in the second frame.
Two-dimensional input is drawn at z = 0, initially face-on. The three vertex
colors are fixed across frames. No labels or categories are inferred from them.
The player's text labels describe frames, not observations.

An all-missing row means inactive, preserving its identity for later frames.
A partially missing row is invalid. Bounds stay fixed across all original
frames, and every retained frame has equal duration. No interpolation or
alignment occurs. Record meaningful original frame labels for a real solver
trace or measured sequence; uniform playback does not recover irregular time
intervals. The [animation guide](animation.html) gives larger examples and
explicit frame-selection recipes.

### Save and clean up

The example below checks a small HTML export and, when magick is installed,
a three-frame GIF, then removes all outputs. It does not launch a viewer.
Choose your own persistent directory when keeping an export.

```{r export-recipe, eval=have.rgl}
local({
  directory <- tempfile("ivue-recipe-")
  dir.create(directory)
  on.exit(unlink(directory, recursive = TRUE))
  htmlwidgets::saveWidget(player, file.path(directory, "triangle.html"),
                          selfcontained = FALSE)
  stopifnot(file.exists(file.path(directory, "triangle.html")))
  if (requireNamespace("magick", quietly = TRUE)) {
    gif <- write.animation.gif(player, file.path(directory, "triangle.gif"),
                               width = 240, height = 240, final.hold = 0)
    stopifnot(file.exists(gif))
  }
})
```

HTML retains interaction and must accompany its dependency folder unless
saved with `selfcontained = TRUE` (which needs Pandoc). GIF is a fixed
orthographic raster rendering using the R widget's initial camera; it does
not capture browser rotations. See the [export comparison](function-guide.html#recorded-coordinate-animation-and-export)
for controls, portability, and limitations.

## Inspect the retinal bundle without refitting

The bundled example adds a real identity and provenance check. Rendering the
full case study is left to the [retinal-development vignette](retinal-development.html),
which includes static posters and complete drawing recipes.

```{r retinal-structure}
retina <- readRDS(system.file("extdata", "retinal-development.rds", package = "ivue"))
ids <- retina$annotations$id
stopifnot(identical(names(retina$coordinates), c("umap", "sknn")),
          identical(dim(retina$annotations), c(12000L, 3L)),
          !anyNA(ids), !anyDuplicated(ids),
          identical(retina$graph$vertices, ids),
          nrow(retina$graph$edges) == 4108L,
          all(retina$graph$edges$from %in% ids),
          all(retina$graph$edges$to %in% ids))
for (coords in retina$coordinates) {
  stopifnot(identical(dim(coords), c(12000L, 3L)),
            identical(rownames(coords), ids), all(is.finite(coords)))
}
data.frame(representation = names(retina$coordinates),
           observations = 12000L, dimensions = 3L)
c(age.levels = nlevels(retina$annotations$age),
  cell.type.levels = nlevels(retina$annotations$cell.type),
  displayed.edges = nrow(retina$graph$edges))
```

There are two 12,000-by-3 matrices (`umap`, `sknn`), an undirected distance
weighted graph with 4,108 edges, and an annotation table with synthetic `id`,
10-level `age`, and 11-level `cell.type`. `provenance` records the source,
fitting cohort, methods, sample, checksums, and upstream package versions.

The upstream fitting cohort had **120,804 cells**. The published analysis
retained **107,052 retinal cells**; the bundle displays a stratified sample
of **12,000** retained cells. Both embeddings were fitted on the full cohort
before extracting these display rows. The displayed graph is an induced
subgraph of the full fitting graph: it keeps only edges whose endpoints both
survive display selection and can be disconnected. No embedding is refitted
on this smaller graph.

Published UMAP used **Canberra distance** on the upstream representation.
The symmetric 4-nearest-neighbor graph used **Euclidean distance**, followed
by weighted-GRIP and edge-KK layout. Distance metrics define comparisons of
upstream observations; embedding/layout methods produce display coordinates.
ivue renders those results. UMAP coordinates were centered for bundling;
the README capture additionally applies independent rigid rotations to face
each P14 centroid toward the viewer. The recipe camera alone does not perform
that alignment. The README's PHATE view uses Euclidean distance and was also
fitted on the full cohort, but its coordinates occur only in external assets,
not in this RDS. No raw expression, principal-component matrix, original
barcodes, or sample identifiers are bundled.

The values come from [Clark et al. (2019)](https://doi.org/10.1016/j.neuron.2019.04.010),
GEO GSE118614, and the Goff lab's published coordinates and annotations.
Their [source use agreement](https://github.com/gofflab/developing_mouse_retina_scRNASeq/blob/3bfeea29ecc957d59e4a156229449b060ffda0a8/README.md#use-agreement)
contains a prepublication consent restriction tied to January 1, 2019.
The documented source review found no separate dataset license. Public
availability and the elapsed date are not a public-domain claim or a new
permission grant; the package's GPL code license does not establish rights
in third-party values. The installed `extdata/README.md` and stored
`provenance$source.terms` preserve these limitations.

Shared colors and IDs help compare the same cells, but apparent proximity,
separation, or a smooth age gradient is not by itself evidence of equivalent
geometries, clustering fidelity, lineage, or biological causation. Interpret
the plots together with the upstream analyses and the case study's limitations.
