---
title: "Finding your way around ivue"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Finding your way around ivue}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ivue)
have.rgl <- nzchar(system.file(package = "rgl"))
```

ivue turns supplied coordinates and annotations into interactive views. Start
with what you want to show, then choose an entry point. The
[example data and recipes](example-data.html) guide supplies small, reproducible
inputs. The [introduction](ivue-introduction.html) develops plotting controls,
the [retinal case study](retinal-development.html) explains a real dataset,
and the [animation guide](animation.html) covers longer recorded sequences.

## Choose a starting point

| Your task | Start here | Decision |
|---|---|---|
| Inspect a point cloud | `plot3D.plain(X)` | One color or explicit row colors; no annotation legend. |
| Show a numerical annotation | `plot3D.cont(X, values)` | Continuous or explicitly binned colors with a matching legend. |
| Show categories | `plot3D.groups(X, groups)` | Named groups, which need not be clusters. |
| Inspect an embedded graph | `prepare.graph()` then `plot3D.graph()` | Check IDs and edges before adding coordinates and annotations. |
| Compare views consistently | A reusable color scale and `camera.zup()` | Fix the meaning of colors and the initial orientation. |
| Add geometric context | `layers = list(...)` | Supply connectivity, labels, a mesh, a reference surface, or axes. |
| Play recorded coordinates | `animate.frames()` | Preserve row identity across all frames. |
| Share a view | `htmlwidgets::saveWidget()` or `write.animation.gif()` | Choose interactive HTML or a fixed raster animation. |

## A first view

This illustrative helix uses base R and a numerical annotation: height. No
embedding or model is fitted.

```{r first-data}
t <- seq(0, 2 * pi, length.out = 24)
X <- cbind(x = cos(t), y = sin(t), z = seq(-1, 1, length.out = 24))
rownames(X) <- sprintf("point-%02d", seq_len(nrow(X)))
height <- X[, "z"]
stopifnot(ncol(X) == 3L, all(is.finite(X)), length(height) == nrow(X))
height.scale <- color.scale.cont(height, limits = c(-1, 1))
head(map.colors(height, height.scale)$colors)
```

```{r first-view, eval=have.rgl}
view <- plot3D.cont(X, values = height, scale = height.scale,
                   legend.title = "Height", legend.width = 120,
                   point.size = 7, height = 340,
                   camera = camera.zup(zoom = 0.65))
view
```

The assignment constructs a widget on a private rgl null device. The final
`view` expression displays it in this document; interactive printing uses
RStudio's Viewer or a browser. Construction itself opens neither a browser
nor a native graphics window. Saving is a third operation:

```{r save-first-view, eval=have.rgl}
local({
  directory <- tempfile("ivue-html-")
  dir.create(directory)
  on.exit(unlink(directory, recursive = TRUE))
  htmlwidgets::saveWidget(view, file.path(directory, "view.html"),
                          selfcontained = FALSE)
  stopifnot(file.exists(file.path(directory, "view.html")))
})
```

This checks saving, then removes the temporary HTML and its dependency folder.
Choose a persistent destination to keep your own export. If rgl is absent,
the widget and save chunks are skipped; data preparation and color mapping
still run. Reading an already rendered vignette needs neither rgl nor R,
but interacting with its scene requires a browser with WebGL.

## Keep observations aligned

Static plots require a numeric matrix or all-numeric data frame with **exactly
three columns**, at least one row, and finite coordinates. Each row is an
observation. For a planar static plot, explicitly append `z = 0` to two-column
coordinates. Missing coordinates, `NaN`, and infinity cause errors; rows are
never silently cleaned or dropped.

Named point-cloud `values`, `groups`, per-point colors, style color vectors,
and logical highlight masks match **observation IDs in `rownames(X)`**. Explicit
row names must be unique, nonempty, and nonmissing. Annotation names must match
the complete ID set exactly; partial, duplicate, missing, or extra names fail.
Automatic data-frame row numbers are not observation IDs. Named annotations
without explicit coordinate IDs also fail. Unnamed vectors follow **row position**;
use `unname()` explicitly when their names are not IDs. Only unnamed scalar
colors are recycled. Coordinates keep their original order, and numeric
highlight indices and indexed layers still refer to those row positions.
See [the recipes](example-data.html#construct-a-point-cloud). Missing numerical
annotations use `na.color`; infinite annotation values are errors. Missing
categories also use the missing color. Neither removes coordinate rows.
Plain colors must be valid R colors with no missing entries, of length one
or the number of rows.

For graph plots, vertex IDs are unique, nonempty, nonmissing strings.
Unnamed coordinates and annotations follow `graph$vertices$id` order. Named
coordinate rows, `values`, `groups`, per-vertex `col`, style color vectors,
and logical highlight masks must match that entire ID set exactly and are
reordered to graph order. Partial, duplicate, missing, or extra names fail.
Numeric highlight indices and all indexed layers refer to **graph vertex
order after alignment**, not to the supplied coordinate order. Never reorder
a prepared vertex table without remapping its integer edge endpoints.

Highlighting changes styling, not membership or the fitted scale. Supply a
logical mask without missing entries or one-based row indices; `NULL` selects
all observations. `highlight.style` and `non.highlight.style` override point
type, size, radius, color, or alpha. Style color vectors align to all rows.
The legend describes the base scale with global alpha; it does not describe
highlight overrides. See [groups and highlighting](ivue-introduction.html#groups-and-highlighting).

## Function catalog

This catalog covers all **18 explicit public function exports**, once each.
ivue registers **2 S3 methods**: `print()` summarizes prepared graphs and color
scales without rendering; `$vertices`, `$edges`, and scale fields retain the full
data. These methods are described through their object workflows, not counted
as separate function rows. ivue re-exports no functions. Dots in names such
as `plot3D.cont` do not make them registered methods of `plot3D()` or `plot()`.
Call these functions directly. The returned widgets use printing and knitting
methods supplied by htmlwidgets/rgl; ivue does not add a separate widget
printing interface. Prepared graphs, scales, and layers are lists carrying
classes, inspected with ordinary R list operations.

In an R session, every name below has help, for example
`help("plot3D.cont", package = "ivue")`. Shared help topics describe related
functions together. Internal dot-prefixed helpers are implementation details.
The repository's `make audit-guide` checks catalog coverage and help aliases.

### Point clouds and annotations

| Function | Purpose | Principal input | Returned object |
|---|---|---|---|
| `plot3D.plain()` | Draw supplied point colors. | Finite n-by-3 `X`, scalar or row colors. | rglwidget/htmlwidget. |
| `plot3D.cont()` | Draw numerical colors and a legend. | `X`, one numerical value per row, optional scale. | Widget with mapping metadata. |
| `plot3D.groups()` | Draw categorical colors and a legend. | `X`, one group per row, optional scale. | Widget with mapping metadata. |

`attr(widget, "ivue")` records coordinates, integer `row.ids`, explicit
`observation.ids` (or `NULL` for unnamed coordinates), mapped colors, highlight,
draw IDs, aspect, camera, and the captured scene. Draw IDs describe serialized
scene objects, not a device left open for further drawing.

### Graph preparation and plotting

| Function | Purpose | Principal input | Returned object |
|---|---|---|---|
| `prepare.graph()` | Validate topology and expose vertex/edge order. | Graph table, lists, matrix, or igraph object. | `ivue_graph` list: vertices, edges, directedness, weight type. |
| `plot3D.graph()` | Draw an embedded graph with optional annotations. | Graph plus exactly one of `X` or `layout`. | Widget with prepared graph metadata. |

Accepted formats are an edge data frame with `from`, `to`, `weight` and an
explicit `vertices` argument (including isolates); a list with `edges` and
`vertices`; paired `adj.list`/`weight.list` lists whose neighbors are integer
row indices; dense square adjacency matrices; Matrix sparse adjacency
matrices; and igraph objects. A prepared `ivue_graph` can be reused and is
revalidated. Vertex attributes are retained. Table edges retain input order;
reciprocal undirected list/matrix edges retain one copy from the lower vertex
index. Edge widths and colors follow this prepared edge order.

Matrices use zero for absence; use tables/lists for actual zero-weight edges.
Explicit sparse zeros are rejected. Undirected adjacency must be reciprocal
with equal weights. Self-loops and parallel edges are rejected, and directed
data can be prepared but cannot be rendered in this release.

With supplied `X`, ivue does not compute a layout or construct an igraph
object for table/list/matrix inputs. Built-in `layout = "kk"` and `"fr"`
require igraph: Kamada-Kawai uses positive **distances**, whereas
Fruchterman-Reingold uses positive **strengths**. Declare `weight.type`
accordingly; no inversion is performed. `"unweighted"` permits only unit or
missing weights. Missing weights otherwise fail. Finite zero/negative weights
can be stored and drawn with supplied coordinates, but not used by these
layout adapters. Weights never automatically set visual edge width or color.
A custom `layout` function receives the prepared graph and returns n-by-3
coordinates. ivue provides no general embedding or graph-construction API.
See the [weighted graph workflow](ivue-introduction.html#weighted-graphs-and-geometric-layers).

### Reusable color scales and legends

| Function | Purpose | Principal input | Returned object |
|---|---|---|---|
| `color.scale.cont()` | Fit a continuous or binned numerical scale. | Reference values, optional limits, palette, center or breaks. | `ivue_color_scale` list. |
| `color.scale.groups()` | Fix category ordering and colors. | Reference groups/factor and optional named colors. | `ivue_color_scale` list. |
| `map.colors()` | Apply a fitted scale without plotting. | Values/groups and their scale. | List with row-aligned colors, legend data frame, and scale. |

Continuous scales use the reference range by default; `limits` fixes it.
`center` requires a supplied diverging palette or `color.map` and makes automatic
limits symmetric about it. For a palette, it places the center at the palette
midpoint. A custom `color.map` instead receives data-unit values unchanged by
centering; with explicit limits, `center` does not alter callback colors.
With explicit limits, the center must lie strictly inside them. A constant
range without a center maps to the middle color; all-missing reference values
default to limits `c(0, 1)`.

Out-of-range values are squished to the endpoints by default. `oob = "censor"`
uses `na.color`; `oob = "error"` rejects them. Missing values use `na.color`
(default `"gray80"`). No default winsorization occurs. In `mode = "binned"`,
choose uniform or quantile bins, or explicit increasing breaks; the intervals
are right-closed, with the lowest endpoint included. Explicit limits must
agree with outer breaks. `winsor.p` is only for uniform binned fitting, and
`center` is only for continuous mode. Automatic legend precision distinguishes
boundaries without changing the numerical calculations.

Factors retain level order, including unused nonmissing levels. Other group
vectors use first-occurrence order. When plots fit a default categorical scale,
this is the supplied annotation order before ID alignment, for both point and
graph views. Reuse a scale when annotation order or membership changes.
Named colors must cover the reference
levels; unknown groups error unless `unknown = "missing"`. Missing values are
distinct from a literal `"NA"` category; empty strings are valid groups.
Numeric palette indices are resolved when fitting scales. A `color.map`
function runs at mapping time on data-unit values after out-of-range handling.
It must return one color per value and be deterministic and **pointwise**:
the same value must have the same color when mapped alone, reordered, or in
another batch. Observations, legend ticks, and the ramp are separate calls.
`function(x) ifelse(x < 0, "blue", "red")` obeys that contract; recomputing
`range(x)` or ranks inside each call does not, even without mutable external
state. Prefer a palette with fixed limits for ordinary comparisons. Supply
`color.map` or `palette`, not both.

`map.colors()` returns a legend with `label`, `color`, and `count`. Continuous
ticks have missing counts; binned/category counts describe the mapped input.
A missing entry appears when needed. Reuse one scale across related views:
separately autoscaled panels can give the same color to different numbers.
The [shared-view recipe](example-data.html#share-a-scale-and-camera) checks this
explicitly. Animation accepts the mapped colors through `col`, but does not
add an annotation legend itself.

### Geometric layers

| Function | Purpose | Principal input | Returned object |
|---|---|---|---|
| `layer3D.edges()` | Add straight segments. | Two-column matrix of one-based endpoint indices. | `ivue_layer` specification. |
| `layer3D.path()` | Connect observations in a chosen order. | Ordered one-based row indices. | `ivue_layer` specification. |
| `layer3D.labels()` | Label selected observations. | Row indices and equally many text labels. | `ivue_layer` specification. |
| `layer3D.mesh()` | Add supplied triangular faces. | Three-column matrix of one-based vertex indices. | `ivue_layer` specification. |
| `layer3D.surface()` | Add an independent gridded height surface. | Monotone `x`, `y`, and finite `z` matrix. | `ivue_layer` specification with its own coordinates. |
| `layer3D.axes()` | Add axes intersecting at an origin. | Optional origin, limits, labels, and styles. | `ivue_layer` specification. |
| `layer3D.callback()` | Run advanced rgl drawing before capture. | Function of scene context and named additional arguments. | `ivue_layer` specification. |

Combine specifications with `layers = list(edge.layer, label.layer, ...)` in
a plotting call. Edges, paths, labels, and meshes use plotting rows (aligned
graph rows for graph plots). Label `offset` has three components in coordinate
units. Mesh colors are per face and do not inherit point colors. Meshes retain
supplied connectivity and do not repair folds or degenerate geometry.

A surface has its own grid: `z[i, j]` belongs to `(x[i], y[j])`, so `z` has
`length(x)` rows and `length(y)` columns. Both axes are strictly monotone;
coordinates must be finite. Grid cells split into planar triangles. No
alignment or rescaling occurs. Surface bounds contribute to the scene, but
automatic origin-axis limits use the plotted observations. See the
[introduction](ivue-introduction.html) for meshes and reference surfaces.

`layer3D.axes()` crosses at `c(0, 0, 0)` by default, with positive arrowheads
and no ticks. It is distinct from ordinary bounding-box axes enabled by
`axes = TRUE`; normally use `axes = FALSE` with this layer. Its 3-by-2
`limits` set endpoints, not clipping. The camera is unchanged.

Callbacks run once during scene construction on the private rgl device,
**before serialization**, not on later browser events. Their first argument
contains `X`, integer `row.ids`, `observation.ids`, `colors`, `highlight`, and `draw.ids`; `args` supplies
named additional arguments. They must not open, close, or switch devices.
Captured object IDs are not live devices. Constructing the specification does
not run the callback; drawing its plot does. Prefer ordinary layer constructors
when they express the intended geometry.

### Cameras and aspect ratios

| Function | Purpose | Principal input | Returned object |
|---|---|---|---|
| `camera.zup()` | Specify an initial z-up view. | Elevation, turn, field of view, zoom. | List with 4-by-4 userMatrix, fov, zoom. |

`point.size` measures screen pixels. `point.type = "sphere"` instead uses
`sphere.radius` in coordinate units (default one percent of the largest span,
with a minimum of 1e-8). Setting radius alone does not select spheres.
Spheres carry spatial extent and can overlap; large scenes cost more to draw.
`aspect = "equal"` preserves equal units along all axes. `"normalized"`
stretches axes independently, changing relative distances and proportions.

`fov = 0` is orthographic: distance from the camera does not shrink an object.
Positive field of view introduces perspective foreshortening. Static plots
default to `camera.zup(elevation = 20, turn = -135, fov = 0, zoom = 0.8)`.
Away from the poles, positive z initially projects upward. Interactive rotation
can tilt it; this is not an enforced rotation constraint. Explicit `theta`,
`phi`, or `userMatrix` selects the alternative camera controls described in
`help("plot3D.plain", package = "ivue")`. Browser rotation does not update
R-side camera metadata or synchronize another widget. Open **View controls**
for keyboard rotation/zoom and **Reset view**. **Download view settings** saves
an R recipe containing the current camera, bounds, and aspect. Source it and
supply those settings explicitly to another plot:

```{r recovered-camera, eval=FALSE}
source("ivue-view.R")
recovered <- plot3D.plain(X, camera = view$camera, limits = view$limits,
                          aspect = view$aspect)
```

The file contains viewing settings, not coordinates, annotations, or layers.
Use matching widget dimensions as well for equal screen scale. The camera is
transferable, but a bounds range from one scene may not contain another dataset.
`limits = rbind(c(-2, 2), c(-2, 2), c(-2, 2))` fixes the x/y/z framing range;
all observations must fit. Spheres and layers cannot enlarge it, and geometry
outside the range may fall outside the viewport. It is not a clipping box.
See the [common-bounds recipe](example-data.html#compare-spatial-extents).

### Recorded-coordinate animation and export

| Function | Purpose | Principal input | Returned object |
|---|---|---|---|
| `animate.frames()` | Play recorded positions with timeline/speed controls. | At least two identically shaped n-by-2 or n-by-3 matrices; optional fixed edges. | Widget with player and `ivue.animation` metadata. |
| `write.animation.gif()` | Export retained animation frames. | An animation widget and `.gif` destination. | Normalized output path, invisibly. |

Frame rows keep stable observation identity. If row names are present, every
frame must have the same unique, nonempty names in the same order; frames are
not automatically reordered. Each row must be entirely finite or entirely
missing (`NA`/`NaN`), meaning inactive. Partially missing rows and infinity
fail. Incident edges are hidden while either endpoint is inactive. An empty
frame is allowed, but the whole sequence must contain a finite point. Two
columns are embedded at z = 0 with a face-on default camera; three use z-up.

`labels` means one string per **original frame**, not vertex labels.
`frame.index` selects increasing original indices and overrides `max.frames`
(default 100 evenly spaced frames, retaining endpoints). Bounds use all original
frames, even those omitted. Retained frames have equal playback duration.
These may represent algorithm iterations or physical measurements, but the
player does not infer elapsed time, interpolate, or align coordinates.
Camera rotation changes the view, not the recorded positions. The
[animation guide](animation.html) distinguishes these operations in detail.

| Output | Controls and projection | Dependencies and files |
|---|---|---|
| Interactive HTML | Rotation and animation controls; initial orthographic or perspective view. | `htmlwidgets::saveWidget()` writes HTML. `selfcontained = TRUE` needs Pandoc and embeds dependencies; `FALSE` writes a companion directory that must travel with the HTML. Viewing requires WebGL, not a running R session. |
| GIF | Fixed initial orthographic camera; no interaction. Browser rotations/speed changes are not returned to R. | `write.animation.gif()` needs magick and the R animation object, not saved HTML. Writes one GIF, using a separate raster renderer rather than a WebGL screenshot. |

GIF controls are `fps` (0.1--100), `width`/`height` (64--8192 pixels),
`final.hold` (0--600 additional seconds), `loop`, `labels`, and `overwrite`.
The parent directory must exist and existing files are protected by default.
GIF delays round to centiseconds; points and edges need not look pixel-identical
to WebGL, especially at intersections. Perspective cameras are rejected.
`selfcontained` and `libdir` belong to `htmlwidgets::saveWidget()`, not GIF
export. Neither export operation automatically captures later browser
interaction; download view settings and reconstruct the R widget to reuse it.

## Optional packages and detailed help

Loading ivue does not load rgl or install anything. Color scales, mapping,
cameras, layers, and table/list/dense-matrix graph preparation work without rgl.
Sparse inputs need Matrix; igraph inputs and built-in layouts need igraph.
Constructing static or animation widgets needs rgl. The optional geometry
package can create triangulations in a recipe, but ivue's mesh layer only
needs supplied triangle indices. GRIP traces in the animation guide come
from grip. Shiny integration uses `rgl::rglwidgetOutput()` and
`rgl::renderRglwidget()`; these are not ivue exports.

```{r finding-help, eval=FALSE}
help(package = "ivue")
help("ivue-package", package = "ivue")
help("color.scale.cont", package = "ivue")
vignette("example-data", package = "ivue")
```

The compact task index, short function descriptions, and links to detailed
workflows take inspiration from [Hmisc's documentation](https://hbiostat.org/r/hmisc/)
and its [overview/help organization](https://CRAN.R-project.org/package=Hmisc).
The categories here follow ivue's smaller visualization API.
