---
title: "Point Clouds and Weighted Graphs with ivue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Point Clouds and Weighted Graphs with ivue}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(ivue)
have.rgl <- nzchar(system.file(package = "rgl"))
have.geometry <- requireNamespace("geometry", quietly = TRUE)
```

## Installation and output

`ivue` returns interactive browser widgets. Plotting requires the separately
installed `rgl` package; color scales, mapping, and `prepare.graph()` do not.
The two-package CRAN installation command is `install.packages(c("ivue", "rgl"))`.
For development builds, install ivue from its repository and install
`rgl` separately. Printing a widget displays it in RStudio's Viewer or an
interactive R console's browser. Assigning it to a variable defers viewing.
Optional graph layouts and igraph inputs require
`igraph`, and sparse matrix inputs require `Matrix`.

The plotting functions initialize a private null device before loading `rgl`.
They do not open a native graphics window, save files, or launch a browser.
The private device is closed after scene capture, including on errors, and
caller options and any surviving previous device are restored. An `rgl`
namespace that another package already initialized unsuccessfully may require
a fresh R session.

## A sample from a saddle

The input is an `n x 3` numeric matrix, or an all-numeric data frame. Each row
is one observation. Coordinates must be finite; ivue does not silently drop
rows with missing coordinates.

```{r saddle-data}
set.seed(1)
xs <- runif(250, -1, 1)
ys <- runif(250, -1, 1)
zs <- 1.2 * (xs^2 - ys^2)
X <- cbind(x = xs, y = ys, z = zs)
```

```{r saddle-plain, eval=have.rgl}
plain <- plot3D.plain(X, col = "#197A68", point.size = 5,
                     axes = TRUE, xlab = "x", ylab = "y", zlab = "z")
plain
```

`point.size` is a screen-pixel size. Spheres are a separate primitive, with
`point.type = "sphere"` and `sphere.radius` measured in coordinate units.
For example, `plot3D.plain(X, point.type = "sphere", sphere.radius = 0.02)`
draws spheres of radius 0.02. A missing radius uses one percent of the largest
coordinate span. Dense sphere scenes can be much more expensive than points.

The default `aspect = "equal"` preserves equal units across axes. Choose
`aspect = "normalized"` only when stretching axes independently is intended;
it changes relative geometric distances.

## Reusable numerical colors

Continuous interpolation is the default. Here zero has a scientific meaning,
so we specify both a center and a diverging palette. Automatic limits become
symmetric around that center.

```{r height-scale}
height.scale <- color.scale.cont(zs, center = 0,
  palette = c("#2166AC", "#F7F7F7", "#B2182B"))
mapping <- map.colors(zs, height.scale)
head(mapping$colors)
mapping$legend
```

```{r saddle-continuous, eval=have.rgl}
colored <- plot3D.cont(X, values = zs, scale = height.scale,
                       point.size = 5, legend.title = "Saddle height")
colored
```

Construct one scale from a reference sample and reuse it to compare plots.
`limits` fixes the numerical range. Values outside it are squished by default;
`oob = "censor"` maps them to the missing color, while `oob = "error"` rejects
them. Missing values use `na.color` without removing coordinate rows.

For binned colors, request them explicitly. Break calculations retain full
precision. The default `digits = NULL` increases legend precision until nearby
boundaries are distinguishable; an explicit `digits` fixes precision and warns
if labels are ambiguous. There is no automatic
winsorization. `center` is a continuous-scale control; for binned diverging
colors, supply the intended breaks and bin colors directly.

```{r binned-scale}
binned <- color.scale.cont(zs, mode = "binned",
  breaks = c(-1.2, -0.4, 0.4, 1.2),
  palette = c("#2166AC", "#EEEEEE", "#B2182B"))
map.colors(zs, binned)$legend
```

`palette` can be a vector of colors or a function of the requested number of
colors. `color.map` instead receives numerical values and must return one
color per value. Do not supply both. Numeric palette indices, including missing
colors, are fixed when a scale is fitted, so later changes to R's `palette()`
do not change its colors. A `color.map` callback is run at mapping time; its
dependence on mutable external state remains the caller's responsibility.

## Textbook-style axes and a z-up camera

The earlier saddle sample is uniform in its horizontal parameter square, not
in surface area. For `z = C * (x^2 - y^2)`, the surface-area factor is
`sqrt(1 + 4*C^2*(x^2 + y^2))`. Accepting uniform-square proposals with probability
proportional to this factor gives a surface-area-uniform sample. Batches
accumulate until there are at least `n` accepted points; the final line retains
exactly `n` of them.

```{r area-uniform-saddle}
set.seed(1)
n <- 500
C <- 0.8
half.width <- 1
xy <- matrix(numeric(), ncol = 2)
while (nrow(xy) < n) {
  proposal <- matrix(runif(4000, -half.width, half.width), ncol = 2)
  area <- sqrt(1 + 4*C^2 * rowSums(proposal^2))
  accept <- runif(nrow(proposal)) < area / sqrt(1 + 8*C^2*half.width^2)
  xy <- rbind(xy, proposal[accept, , drop = FALSE])
}
xy <- xy[seq_len(n), , drop = FALSE]
saddle <- cbind(x = xy[, 1], y = xy[, 2],
                z = C * (xy[, 1]^2 - xy[, 2]^2))
byr.scale <- color.scale.cont(saddle[, "z"], center = 0,
                               palette = c("blue", "yellow", "red"))
```

`layer3D.axes()` adds three axes through the origin, with arrowheads at their
positive ends. It does not rotate the data or choose a camera. Pair it with
`camera.zup()` to put z vertically upward in the initial view:

```{r saddle-coordinate-axes, eval=have.rgl}
saddle.view <- plot3D.cont(
  saddle, values = saddle[, "z"], scale = byr.scale,
  point.type = "sphere", sphere.radius = 0.02,
  legend.title = "Saddle height", legend.width = 160,
  axes = FALSE, aspect = "equal",
  layers = list(layer3D.axes(head.length = 0.04, head.angle = pi/8)),
  camera = camera.zup(elevation = 20, turn = -135, fov = 0, zoom = 0.4),
  height = 500L
)
```

```{r saddle-coordinate-display, eval=have.rgl, echo=FALSE}
# Keep the enlarged scene in a 7:5 frame when the vignette column narrows.
# Move the legend above the frame on small screens rather than hiding data.
saddle.view$elementId <- "saddle-coordinate-example"
htmltools::tagList(
  htmltools::tags$style(htmltools::HTML("
    #saddle-coordinate-example, #saddle-mesh-example {
      height: auto !important; aspect-ratio: 7 / 5;
    }
    @media (max-width: 600px) {
      #saddle-coordinate-example, #saddle-mesh-example { margin-top: 160px; }
      #saddle-coordinate-example > .ivue-legend,
      #saddle-mesh-example > .ivue-legend {
        top: -155px !important; max-height: none !important;
      }
    }
  ")),
  saddle.view
)
```

Zero is yellow; negative and positive heights move toward blue and red,
respectively. `scale` takes a color-scale object, not a raw color vector.
Reusing `byr.scale` and the original height values preserves observation colors
if these points are subsequently displayed in another coordinate configuration.

The axes extend beyond the data automatically. Set `origin` to move their
intersection, or use explicit `limits`, for example
`rbind(x = c(-1.2, 1.2), y = c(-1.2, 1.2), z = c(-1, 1))`, to fix endpoints
across views. These limits do not clip data. `head.length` is a fraction of the
full axis span; `head.angle` is the cone half-angle in radians. `width`, `col`,
`cex`, and `label.offset` control the shafts, colors, label size, and label gaps.
The heads are solid cones in data coordinates, so they keep their 3D geometry
while the widget rotates. Use equal aspect to preserve their proportions.

With `turn = -135`, positive x points down-left and positive y down-right.
With `turn = 0`, positive x is horizontal to the right and positive y recedes.
`elevation` changes the angle above the xy plane; `fov = 0` removes perspective.
The helper only specifies the initial view: dragging can still tilt z, and at
elevations of plus or minus 90 degrees z points along the viewing direction.
Neither the axes constructor nor the camera helper needs rgl until a plot is
actually drawn. The same layer and camera also work with `plot3D.graph()`.

## A triangular mesh beneath the spheres

Connecting neighboring observations with triangular faces makes the saddle's
shape easier to see. Construct a Delaunay triangulation in the original
two-dimensional parameter plane, then draw its faces at the three-dimensional
saddle coordinates. Triangulating the 3D coordinates instead would construct
a different object, not this surface mesh.

The `geometry` package is an add-on, not part of base R. Install it with
`install.packages("geometry")` to run this example. It is suggested for the
vignette, but `layer3D.mesh()` does not require it when triangle indices are
already available. The example below is evaluated when both `geometry` and
`rgl` are installed.

```{r saddle-triangulation, eval=have.geometry}
triangles <- geometry::delaunayn(saddle[, c("x", "y")])
surface <- layer3D.mesh(
  triangles, col = "gray75", alpha = 0.2,
  edge.col = "gray45", edge.alpha = 0.35, edge.width = 1
)
```

Each row of `triangles` contains three indices into `saddle`. The mesh joins
those observations with planar faces, approximating the surface over the
sample's parameter-space convex hull. It does not extend to unsampled corners
of the square. The points, colors, axes, and initial camera below are the same
as in the preceding plot.

```{r saddle-mesh, eval=have.rgl && have.geometry}
saddle.mesh <- plot3D.cont(
  saddle, values = saddle[, "z"], scale = byr.scale,
  point.type = "sphere", sphere.radius = 0.02,
  legend.title = "Saddle height", legend.width = 160,
  axes = FALSE, aspect = "equal",
  layers = list(surface, layer3D.axes()),
  camera = camera.zup(elevation = 20, turn = -135, fov = 0, zoom = 0.4),
  height = 500L
)
```

```{r saddle-mesh-display, eval=have.rgl && have.geometry, echo=FALSE}
saddle.mesh$elementId <- "saddle-mesh-example"
saddle.mesh
```

`alpha` controls face opacity; `edge.alpha` independently controls the mesh
lines. Shared edges are drawn once. Set `alpha = 0` for a wireframe or
`edges = FALSE` to show faces without their outlines.

To inspect another embedding of these same observations, retain `surface`
and replace only the plotting coordinates. For example, if `Z` contains the
metric-MDS or metric-MDS-followed-by-edge-KK coordinates in the same row order:

```{r mesh-other-embedding, eval=FALSE}
plot3D.cont(
  Z, values = saddle[, "z"], scale = byr.scale,
  point.type = "sphere", sphere.radius = 0.02, axes = FALSE,
  layers = list(surface, layer3D.axes()), camera = camera.zup()
)
```

Do not retriangulate `Z`: fixed triangle connectivity reveals how the original
mesh stretches, folds, or collapses. Original heights continue to determine
point colors. The mesh is a visualization overlay, separate from any
symmetric-kNN graph used for fitting or fidelity scoring; adding it does not
change that graph or its objectives.

## Groups and highlighting

Groups need not be clusters. Factors retain their level order; other vectors
use first-occurrence order. Named colors make the assignment explicit and
stable across datasets. Unknown groups raise an error unless the scale uses
`unknown = "missing"`. Missing factor levels are treated as missing, not as
the literal group `"NA"`. Empty strings remain valid groups. Legends quote
group labels when needed to distinguish an empty label or a group called
`"Missing"` from the missing-value entry.

```{r group-scale}
groups <- factor(ifelse(zs < 0, "Negative", "Nonnegative"),
                 levels = c("Negative", "Nonnegative"))
group.scale <- color.scale.groups(groups,
  colors = c(Negative = "#D95479", Nonnegative = "#009F87"))
map.colors(groups, group.scale)$legend
```

```{r saddle-groups, eval=have.rgl}
plot3D.groups(X, groups, scale = group.scale, point.size = 5,
  highlight = abs(zs) > 0.4,
  highlight.style = list(point.size = 7),
  non.highlight.style = list(col = "gray75", alpha = 0.25))
```

`highlight` changes styling, not membership: all 250 rows keep their original
IDs and the color scale is not refitted. Legends describe the base color scale
and global `alpha`, not the per-highlight overrides. Scale colors can include
alpha, and global `alpha` multiplies it. A highlight-style `alpha` overrides
the global multiplier for that subset, while preserving its color's alpha.

## Weighted graphs and geometric layers

A graph supplies topology and weights. Coordinates supply geometry; one does
not substitute for the other. The following graph contains an isolated vertex.

```{r graph-data}
vertices <- data.frame(id = c("A", "B", "C", "D"),
                       label = c("Start", "Middle", "End", "Isolate"))
edges <- data.frame(from = c("A", "B"), to = c("B", "C"), weight = c(2, 4))
coords <- rbind(A = c(0, 0, 0), B = c(1, 1, 0),
                C = c(2, 0, 1), D = c(0, 2, 1))
graph <- prepare.graph(edges, vertices = vertices, weight.type = "distance")
graph$edges
coords <- coords[c("C", "A", "D", "B"), ]
values <- c(D = 40, B = 20, A = 10, C = 30)
```

```{r embedded-graph, eval=have.rgl}
plot3D.graph(graph, X = coords, values = values, edge.col = "gray50",
  edge.width = 1 + graph$edges$weight,
  point.type = "sphere", sphere.radius = 0.07,
  layers = list(layer3D.path(c(1, 2, 3), col = "#D55E00", width = 4),
                layer3D.labels(1:4, vertices$label, offset = c(0, 0, 0.15))))
```

The other accepted formats are paired `adj.list`/`weight.list` lists, dense or
sparse adjacency matrices, and igraph objects. For the same graph:

```{r adjacency-list}
adjacency <- list(
  adj.list = list(A = 2L, B = c(1L, 3L), C = 2L, D = integer()),
  weight.list = list(A = 2, B = c(2, 4), C = 4, D = numeric()))
```

Undirected adjacency entries must be reciprocal, with matching weights.
Matrices use zero to mean no edge; use tables or lists for actual zero-weight
edges. Coordinate row names are aligned to vertex IDs. Without row names,
coordinates follow vertex order. Named values and groups are independently
aligned by exact vertex ID, as in the shuffled example above. Unnamed annotation
vectors follow vertex order. Duplicate, missing, partial, and extra names are
rejected. Named point colors, highlight-style color vectors, and logical
highlight masks use the same rule. Layer row indices and numeric highlight
indices refer to graph vertex order after alignment.

Weights do not automatically control edge width or color. The example chooses
`1 + graph$edges$weight` explicitly. Edge colors and widths follow the edge order
exposed by `prepare.graph()` before rendering; matrix/list inputs use one copy of each
undirected edge in vertex-row order. Table inputs retain their supplied order.

Supplying `X` does not construct an igraph object or compute a layout for table,
list, or matrix inputs. To compute geometry instead, supply `layout`, not `X`:

```{r optional-layout, eval=have.rgl && requireNamespace("igraph", quietly=TRUE)}
layout.widget <- plot3D.graph(adjacency, layout = "kk", weight.type = "distance",
                              seed = 1)
stopifnot(inherits(layout.widget, "htmlwidget"))
unweighted <- igraph::make_ring(4)
unweighted.widget <- plot3D.graph(unweighted, layout = "fr", weight.type = "unweighted")
stopifnot(inherits(unweighted.widget, "htmlwidget"))
```

`kk` treats positive weights as distances; `fr` requires positive strengths.
No weight inversion occurs. `weight.type = "unweighted"` allows missing or unit
weights only. Zero and negative weights may be drawn with supplied coordinates,
but cannot be used by these layout adapters. Directed rendering, self-loops,
and parallel edges are explicitly unsupported in this release.

Use `layer3D.edges()` when you already have geometric endpoint pairs, without
requiring a graph object. `layer3D.callback()` supports advanced rgl drawing;
callbacks receive coordinates and row/draw IDs but must not open, close, or
switch devices. Layers are applied before serialization, not appended to a
live browser widget later.

## Cameras, metadata, and export

The widget's metadata describes the captured R scene. Reusing its initial camera
keeps related plots comparable. Rotating a widget in the browser does not update
this R-side camera metadata.

```{r camera-export, eval=have.rgl}
camera <- attr(colored, "ivue")$camera
comparison <- plot3D.plain(X, camera = camera)
stopifnot(identical(attr(comparison, "ivue")$row.ids, seq_len(nrow(X))))

out <- tempfile(fileext = ".html")
htmlwidgets::saveWidget(colored, out, selfcontained = FALSE)
stopifnot(file.exists(out))
unlink(c(out, sub("\\.html$", "_files", out)), recursive = TRUE)
```

For a single portable HTML file, use `selfcontained = TRUE`; Pandoc must be
installed. Otherwise distribute the HTML file together with its dependency
directory. Widgets can also be used in Shiny through `rgl::rglwidgetOutput()`
and `rgl::renderRglwidget()`; Shiny is not an ivue dependency.

Default widget width follows its container, with a height of 600 pixels.
Explicit `width` and `height` set dimensions. Long legends scroll within their
allocated area, and `legend.show = FALSE` permits a separate application legend
built from the same `map.colors()` result.
