---
title: "Point Clouds and Weighted Graphs with ivue"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Point Clouds and Weighted Graphs with ivue}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(ivue)
have.rgl <- nzchar(system.file(package = "rgl"))
```

## Installation and output

`ivue` returns interactive browser widgets. Plotting requires the separately
installed `rgl` package; color scales, mapping, and `prepare.graph()` do not.
The two-package CRAN installation command is `install.packages(c("ivue", "rgl"))`;
for a pre-CRAN development build, install ivue from its repository and install
`rgl` separately. Printing a widget displays it in RStudio's Viewer or an
interactive R console's browser. Assigning it to a variable defers viewing.
Optional graph layouts and igraph inputs require
`igraph`, and sparse matrix inputs require `Matrix`.

The plotting functions initialize a private null device before loading `rgl`.
They do not open a native graphics window, save files, or launch a browser.
The private device is closed after scene capture, including on errors, and
caller options and any surviving previous device are restored. An `rgl`
namespace that another package already initialized unsuccessfully may require
a fresh R session.

## A sample from a saddle

The input is an `n x 3` numeric matrix, or an all-numeric data frame. Each row
is one observation. Coordinates must be finite; ivue does not silently drop
rows with missing coordinates.

```{r saddle-data}
set.seed(1)
xs <- runif(250, -1, 1)
ys <- runif(250, -1, 1)
zs <- 1.2 * (xs^2 - ys^2)
X <- cbind(x = xs, y = ys, z = zs)
```

```{r saddle-plain, eval=have.rgl}
plain <- plot3D.plain(X, col = "#197A68", point.size = 5,
                     axes = TRUE, xlab = "x", ylab = "y", zlab = "z")
plain
```

`point.size` is a screen-pixel size. Spheres are a separate primitive, with
`point.type = "sphere"` and `sphere.radius` measured in coordinate units.
For example, `plot3D.plain(X, point.type = "sphere", sphere.radius = 0.02)`
draws spheres of radius 0.02. A missing radius uses one percent of the largest
coordinate span. Dense sphere scenes can be much more expensive than points.

The default `aspect = "equal"` preserves equal units across axes. Choose
`aspect = "normalized"` only when stretching axes independently is intended;
it changes relative geometric distances.

## Reusable numerical colors

Continuous interpolation is the default. Here zero has a scientific meaning,
so we specify both a center and a diverging palette. Automatic limits become
symmetric around that center.

```{r height-scale}
height.scale <- color.scale.cont(zs, center = 0,
  palette = c("#2166AC", "#F7F7F7", "#B2182B"))
mapping <- map.colors(zs, height.scale)
head(mapping$colors)
mapping$legend
```

```{r saddle-continuous, eval=have.rgl}
colored <- plot3D.cont(X, values = zs, scale = height.scale,
                       point.size = 5, legend.title = "Saddle height")
colored
```

Construct one scale from a reference sample and reuse it to compare plots.
`limits` fixes the numerical range. Values outside it are squished by default;
`oob = "censor"` maps them to the missing color, while `oob = "error"` rejects
them. Missing values use `na.color` without removing coordinate rows.

For binned colors, request them explicitly. Break calculations retain full
precision. The default `digits = NULL` increases legend precision until nearby
boundaries are distinguishable; an explicit `digits` fixes precision and warns
if labels are ambiguous. There is no automatic
winsorization. `center` is a continuous-scale control; for binned diverging
colors, supply the intended breaks and bin colors directly.

```{r binned-scale}
binned <- color.scale.cont(zs, mode = "binned",
  breaks = c(-1.2, -0.4, 0.4, 1.2),
  palette = c("#2166AC", "#EEEEEE", "#B2182B"))
map.colors(zs, binned)$legend
```

`palette` can be a vector of colors or a function of the requested number of
colors. `color.map` instead receives numerical values and must return one
color per value. Do not supply both. Numeric palette indices, including missing
colors, are fixed when a scale is fitted, so later changes to R's `palette()`
do not change its colors. A `color.map` callback is run at mapping time; its
dependence on mutable external state remains the caller's responsibility.

## Groups and highlighting

Groups need not be clusters. Factors retain their level order; other vectors
use first-occurrence order. Named colors make the assignment explicit and
stable across datasets. Unknown groups raise an error unless the scale uses
`unknown = "missing"`. Missing factor levels are treated as missing, not as
the literal group `"NA"`. Empty strings remain valid groups. Legends quote
group labels when needed to distinguish an empty label or a group called
`"Missing"` from the missing-value entry.

```{r group-scale}
groups <- factor(ifelse(zs < 0, "Negative", "Nonnegative"),
                 levels = c("Negative", "Nonnegative"))
group.scale <- color.scale.groups(groups,
  colors = c(Negative = "#D95479", Nonnegative = "#009F87"))
map.colors(groups, group.scale)$legend
```

```{r saddle-groups, eval=have.rgl}
plot3D.groups(X, groups, scale = group.scale, point.size = 5,
  highlight = abs(zs) > 0.4,
  highlight.style = list(point.size = 7),
  non.highlight.style = list(col = "gray75", alpha = 0.25))
```

`highlight` changes styling, not membership: all 250 rows keep their original
IDs and the color scale is not refitted. Legends describe the base color scale
and global `alpha`, not the per-highlight overrides. Scale colors can include
alpha, and global `alpha` multiplies it. A highlight-style `alpha` overrides
the global multiplier for that subset, while preserving its color's alpha.

## Weighted graphs and geometric layers

A graph supplies topology and weights. Coordinates supply geometry; one does
not substitute for the other. The following graph contains an isolated vertex.

```{r graph-data}
vertices <- data.frame(id = c("A", "B", "C", "D"),
                       label = c("Start", "Middle", "End", "Isolate"))
edges <- data.frame(from = c("A", "B"), to = c("B", "C"), weight = c(2, 4))
coords <- rbind(A = c(0, 0, 0), B = c(1, 1, 0),
                C = c(2, 0, 1), D = c(0, 2, 1))
graph <- prepare.graph(edges, vertices = vertices, weight.type = "distance")
graph$edges
coords <- coords[c("C", "A", "D", "B"), ]
values <- c(D = 40, B = 20, A = 10, C = 30)
```

```{r embedded-graph, eval=have.rgl}
plot3D.graph(graph, X = coords, values = values, edge.col = "gray50",
  edge.width = 1 + graph$edges$weight,
  point.type = "sphere", sphere.radius = 0.07,
  layers = list(layer3D.path(c(1, 2, 3), col = "#D55E00", width = 4),
                layer3D.labels(1:4, vertices$label, offset = c(0, 0, 0.15))))
```

The other accepted formats are paired `adj.list`/`weight.list` lists, dense or
sparse adjacency matrices, and igraph objects. For the same graph:

```{r adjacency-list}
adjacency <- list(
  adj.list = list(A = 2L, B = c(1L, 3L), C = 2L, D = integer()),
  weight.list = list(A = 2, B = c(2, 4), C = 4, D = numeric()))
```

Undirected adjacency entries must be reciprocal, with matching weights.
Matrices use zero to mean no edge; use tables or lists for actual zero-weight
edges. Coordinate row names are aligned to vertex IDs. Without row names,
coordinates follow vertex order. Named values and groups are independently
aligned by exact vertex ID, as in the shuffled example above. Unnamed annotation
vectors follow vertex order. Duplicate, missing, partial, and extra names are
rejected. Named point colors, highlight-style color vectors, and logical
highlight masks use the same rule. Layer row indices and numeric highlight
indices refer to graph vertex order after alignment.

Weights do not automatically control edge width or color. The example chooses
`1 + graph$edges$weight` explicitly. Edge colors and widths follow the edge order
exposed by `prepare.graph()` before rendering; matrix/list inputs use one copy of each
undirected edge in vertex-row order. Table inputs retain their supplied order.

Supplying `X` does not construct an igraph object or compute a layout for table,
list, or matrix inputs. To compute geometry instead, supply `layout`, not `X`:

```{r optional-layout, eval=have.rgl && requireNamespace("igraph", quietly=TRUE)}
layout.widget <- plot3D.graph(adjacency, layout = "kk", weight.type = "distance",
                              seed = 1)
stopifnot(inherits(layout.widget, "htmlwidget"))
unweighted <- igraph::make_ring(4)
unweighted.widget <- plot3D.graph(unweighted, layout = "fr", weight.type = "unweighted")
stopifnot(inherits(unweighted.widget, "htmlwidget"))
```

`kk` treats positive weights as distances; `fr` requires positive strengths.
No weight inversion occurs. `weight.type = "unweighted"` allows missing or unit
weights only. Zero and negative weights may be drawn with supplied coordinates,
but cannot be used by these layout adapters. Directed rendering, self-loops,
and parallel edges are explicitly unsupported in this release.

Use `layer3D.edges()` when you already have geometric endpoint pairs, without
requiring a graph object. `layer3D.callback()` supports advanced rgl drawing;
callbacks receive coordinates and row/draw IDs but must not open, close, or
switch devices. Layers are applied before serialization, not appended to a
live browser widget later.

## Cameras, metadata, and export

The widget's metadata describes the captured R scene. Reusing its initial camera
keeps related plots comparable. Rotating a widget in the browser does not update
this R-side camera metadata.

```{r camera-export, eval=have.rgl}
camera <- attr(colored, "ivue")$camera
comparison <- plot3D.plain(X, camera = camera)
stopifnot(identical(attr(comparison, "ivue")$row.ids, seq_len(nrow(X))))

out <- tempfile(fileext = ".html")
htmlwidgets::saveWidget(colored, out, selfcontained = FALSE)
stopifnot(file.exists(out))
unlink(c(out, sub("\\.html$", "_files", out)), recursive = TRUE)
```

For a single portable HTML file, use `selfcontained = TRUE`; Pandoc must be
installed. Otherwise distribute the HTML file together with its dependency
directory. Widgets can also be used in Shiny through `rgl::rglwidgetOutput()`
and `rgl::renderRglwidget()`; Shiny is not an ivue dependency.

Default widget width follows its container, with a height of 600 pixels.
Explicit `width` and `height` set dimensions. Long legends scroll within their
allocated area, and `legend.show = FALSE` permits a separate application legend
built from the same `map.colors()` result.
